Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer

### 2019-01-14 Paper/Paper.tex
References and other edits
Began adding refs and creating a bibtex file. Other light edits, too.
Still need to find out how to compile the bibtex file correctly.

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. Ref  [examples] 
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context. Ref
Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.

@@

Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells. EMT has an important yet complicated role in tumor initiation, progression, and metastasis. Ref 
As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer. Ref
Also, due to their motility, some have viewed them as a natural precursor to metastasis. Ref

Herein, we explore the interconnectedness of these three key systems: the tumor, the immune system, and EMT.
In addition to the interactions already discussed, the immune system affects and is affected by EMT.
The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. Ref 
In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). Ref 
Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
Each of these three pieces will be briefly explored in turn for how they will influence one another.

@@

\subsection{EMT}\label{EMT}
There are many characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
The first is that mesenchymal cells are less susceptible to immune clearance. Ref 
The second is that mesenchymal cells proliferate at a slower rate. Ref

The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. Ref 
However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance. Ref
In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
This quality of mesenchymal cells is then quantified in our model by the parameter, MIE, for mesenchymal immune evasion.

@@

In the following sections of the paper we first introduce the model modularly, which is appropriate diven its multiscale and hybrid nature. We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis. We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.

@@

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model. In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored. Inset shows major processes of immune system regulation on the tissue cells.
B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
C. Example survival curve for one cohort of patients with parameters XXXXX.
D. Global sensitivity analysis of XXX parameters using XXXXX.}
\label{fig:ModelIntro}
\end{figure}

\subsection{Exploring the Model}\label{ExplModel}
In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
If any component [???] of a cell has a pathway mutation, that cell is then counted as a mutated cell and contributes towards the total number of cancer cells.
The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
In the model the EMT spectrum is continuous. While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum. Ref [Tian and Levine lab] 
In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.

@@

For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541. [translate into days?]
Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
Also of interest, as the patient's cells are all rapidly mutating right around this time, most of them are also transitioning to a mesenchymal phenotype with all of them being in a mesenchymal phenotype for a short time during the time period that cancer onset occurs.

>>>>

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. (Restifo 2012)  [examples] 
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context. (De Visser 2006)
Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.

@@

Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
EMT has an important yet complicated role in tumor initiation, progression, and metastasis. (Nieto 2016) % References more specific to this claim might be in this paper's references. 
As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer. (Nieto 2016) % References more specific to this claim might be in this paper's references.
Also, due to their motility, some have viewed them as a natural precursor to metastasis. (Alderton 2012)

Herein, we explore the interconnectedness of these three key systems: the tumor, the immune system, and EMT.
In addition to the interactions already discussed, the immune system affects and is affected by EMT.
The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. (Terry 2017) 
In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). (Terry 2017) 
Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
Each of these three pieces will be briefly explored in turn for how they will influence one another.

@@

\subsection{EMT}\label{EMT}
There are many characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
The first is that mesenchymal cells are less susceptible to immune clearance. (Terry 2017) 
The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.

The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. (Terry 2017) 
However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance. (Terry 2017) % 
In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
This quality of mesenchymal cells is then quantified in our model by the parameter, MIE, for mesenchymal immune evasion.

@@

In the following sections of the paper we first introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.

@@

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model. In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored. Inset shows major processes of immune system regulation on the tissue cells.
B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
C. Example survival curve for one cohort of patients with parameters XXXXX. % I should list out all parameter values?
D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
\label{fig:ModelIntro}
\end{figure}

\subsection{Exploring the Model}\label{ExplModel}
In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
In the model the EMT spectrum is continuous.
While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum. (Hong 2015)
In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.

@@

For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. % translate to days like this?
At the same time as mutated cells are taking over, most cells are transitioning to  a mesenchymal phenotype.
After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.% Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
%We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
